---
title: "R Notebook"
output: html_notebook
---

New notebook to clean up pangene analysis of expression data

Load libraries

```{r}
library(DESeq2)
library(tidyverse)
library(lme4)
library(lmerTest)
```
Functions

```{r}
# Function to calculate tau
calculate_tau <- function(counts) {
  max_count <- max(counts, na.rm = TRUE)
  if (max_count == 0) {
    return(NA)  # Return NA if all counts are 0 to avoid division by 0
  }
  tau <- sum(1 - (counts / max_count), na.rm = TRUE) / (length(counts) - 1)
  return(tau)
}
```

Load data. This is complicated because the way pangene is run, we have one rna.id per gene_name, but the gene_name != gene_name from kallisto. The rna.id is one of the tgut or ggal depending on which mapped better. 

However, this can mean that sometimes a gene duplication in the reference (e.g., an rna.id that maps to two xloc_id), is only represented by one pangene mapping. So, let's start by merging and doing some QC on the duplicates. 


```{r}

# read the gene info table into R, with the tcons/xloc/rna.id mapping

gene_info<-read_delim("merged_TOGA_tgutRef_noDup.tsv", delim="\t", 
             col_names = c("tcons_id", "xloc_id", "galGal_id", "tGut_id", "galGal_toga", "tGut_toga", "equiv_class", "gene_name"))

# convert to something that can be merged with the pangene data 

tcons_pg<-gene_info %>% select(xloc_id, tcons_id, galGal_id, tGut_id) %>% 
  mutate(ggal = gsub("_gGal", "", galGal_id), tgut = gsub("_tGut","",  tGut_id)) %>%
  select(tcons_id, xloc_id, ggal, tgut) %>%
  pivot_longer(cols = ggal:tgut, names_to = "species", values_to = "rna_id")

# pangene info, giving the RNA.id for each gene name in the pangene results table. Only need the XLOC_ID to map to RNA_ID, don't need TCONS_ID

gene_pg_full <- read_delim("pangene_map.txt", delim=":", col_names = c("gene_name", "rna_id")) %>%
  full_join(tcons_pg) %>% filter(!is.na(tcons_id), rna_id != "None", !is.na(gene_name)) 

gene_pg <- gene_pg_full %>% 
  select(xloc_id, gene_name, rna_id, species) %>% distinct()

# chromosomal locations

gene_loc <- read_delim("scrubjay_genes.tsv", delim = "\t",
                     col_names = c("chr", "st", "en", "strand", "id_info")) %>%
  mutate(id_info = str_remove_all(id_info, "\"")) %>%
  separate_wider_delim(id_info, ";", names=c("tcons_id", "xloc_id")) %>%
  mutate(tcons_id = str_remove(tcons_id, "transcript_id ")) %>%
  mutate(xloc_id = str_remove(xloc_id, "gene_id ")) %>%
  mutate(chrom = str_extract(str_remove(chr, "aphWoo1#REF#"), "ch\\d+")) %>%
  select(xloc_id, chrom) %>% distinct()
```
Some QC on gene_pg, which is what we'll use to map the pangene table to the kallisto results, as it has both xloc and gene_name. 


```{r}
gene_pg %>% count(xloc_id) %>% count(num = n)
```

Okay, so there are a number of xloc_id that show up more than once. Let's look at a few.

```{r}
gene_pg %>% add_count(xloc_id) %>% filter(n == 2) %>% arrange(xloc_id)
gene_pg %>% add_count(xloc_id) %>% filter(n > 3) %>% arrange(xloc_id)
```
These mostly look like cases where there are multiple rna_id that map to one XLOC, basically N:1 relationships (gGal:scrubjay). Let's also check to see if there are single genes that map to multiple XLOC_ID.

```{r}
gene_pg %>% add_count(gene_name) %>% count(n)

```
```{r}
gene_pg %>% add_count(gene_name) %>% filter(n ==4) %>% arrange(gene_name)

```

So the cases where one gene_name maps to multiple xloc_ids are likely duplicates in the reference relative to tgut or ggal, e.g. 1:N (ggal:scrubjay), or perhaps split/bad annotations? These are almost all cases where there is some copy number variation, perhaps not surprisingly. 

```{r}
pangene_raw <- read_delim("sj-94nl-a2.pg.CNV.Rtable") %>% 
  pivot_longer(-Gene, names_to = "indiv_haplo", values_to = "count") %>%
  separate_wider_delim(indiv_haplo, delim="#", names=c("indiv", "haplo")) %>%
  select(gene_name = Gene, indiv, count) %>% group_by(gene_name, indiv) %>%
  summarize(count = sum(count))

pangene <- gene_pg %>% add_count(xloc_id, name="xloc_n") %>% add_count(gene_name, name="gene_n") %>% right_join(pangene_raw, relationship = "many-to-many")
```

We seem some many-to-many joins, which are presumably arising from 1:n, n:1, and even n:n mappings. 

First step is to make the pangene data have a single row per xloc_id. So if there is an xloc_id that shows up more than once, we collapse and keep the max count. This prevents duplication when we merge with expression data. 

Before we do that, though, let's just double check the gene_name vs xloc_id counts. 

```{r}
pangene %>% select(gene_name, xloc_id, gene_n, xloc_n) %>% distinct() %>% count(gene_n, xloc_n)
```

Let's look at the ones that are 2:2

```{r}
pangene %>% filter(gene_n == 2, xloc_n == 2) %>% select(gene_name, xloc_id, gene_n, xloc_n) %>% distinct() %>% arrange(xloc_id, gene_name)
pangene %>% filter(gene_name == "POLR1A" | gene_name == "PTCD3") %>% arrange(gene_name)
pangene %>% filter(xloc_id == "XLOC_001196" | xloc_id == "XLOC_000835") %>% arrange(xloc_id)
```
Okay, so looking at one example, we have XLOC_001196 matching to PTCD3 and POLR1A; and also POLR1A matching to XLOC_000835 and XLOC_001196. 

This is a bit complicated. 

When one pangene gene_name maps to multiple XLOC_ID, it is fairly simple, we can just sum the counts across XLOC_ID. 

However, when multiple pangene gene_names map to one XLOC, we need to do something else? The options are either to collapse them into a single xloc, or to split the expression, perhaps by rna_id? I'm not sure what causes this? Maybe we can recover some of these with the cd hit clustering? 

Let's start by working with just the mapping table, without adding counts yet, and try to get a single expression value for each pangene gene_id. 


```{r}
exp_raw<-read_delim("kallisto_COMBINED_abundance.tsv", skip = 1, 
                    col_names=c("tcons_id", "length", "eff_length", "est_counts", 
                                "tpm", "sample", "tissue", "species", "geneID")) %>%
  left_join(gene_info) %>%
  select(xloc_id, tcons_id, tpm, tissue, indiv = sample)
```

This is fine, now we need to add the pangene information (gene_name), via the gene_pg and tcons_pg. For this we actually need the full verions of the gene_pg tibble.


```{r}
exp_raw %>% left_join(gene_pg_full, relationship="many-to-many")
```

Okay, actually, this is kind of a mess. I think if we have N (gene_name) : 1 (xloc), we can't really reasonably split expression, we just have no good way to do this because there is no id we can really track here. 

The simplest thing is to just keep those, with expression duplicated, but mark them to see if our results are robust to filtering on these.

So in that case, we collapse expression into a single TPM per xloc_id, merge to pangene_raw, and then merge with pangene_raw, and then collapse to one observation per gene_name / indiv / tissue for both count and tpm. We need to make some more mapping tables.

```{r}
pangene_mapping <- pangene %>% select(gene_name, xloc_id) %>% distinct()
tpm_by_gene <- exp_raw %>% group_by(xloc_id, indiv, tissue) %>% summarize(tpm = sum(tpm)) %>%
  full_join(pangene_mapping, relationship = "many-to-many") %>% left_join(gene_loc) %>%
  mutate(chrom = replace_na(chrom, "non-aut"))
```

Make the final analysis dataset. This also includes the TPM > 1 filter and log transform.

```{r}
tpm_pg <- tpm_by_gene %>% filter(!is.na(gene_name)) %>% 
  group_by(indiv, tissue, gene_name, chrom) %>% 
  summarize(tpm = sum(tpm)) %>% left_join(pangene_raw) %>%
  filter(!is.na(tpm), !is.na(count)) %>%
  filter(tpm > 1) %>%
  mutate(tpm = log10(tpm))
```

Finally, let's add pangene stats to each gene row, and also add the sex where we have it.


```{r}
ref_count <- pangene_raw %>% filter(indiv == "AW_365336") %>% select(gene_name, ref_count = count) %>% distinct()

pangene_stats <- pangene_raw %>% 
  distinct() %>% 
  separate(indiv, into = c("species", NA), sep = "_", remove=FALSE, extra="merge") %>%
  group_by(gene_name) %>%
  summarize(min_count = min(count), 
            max_count = max(count), 
            AW_var = sum(species == "AW" & count != 2),
            AC_var = sum(species == "AC" & count !=2)) %>%
  left_join(ref_count) %>%
  select(gene_name, ref_count, AW_var, AC_var, min_count, max_count) %>% distinct()
```
```{r}
sex<-read_delim("indiv_sex.tsv", trim_ws = TRUE)
```


```{r}
tpm_pg <- tpm_pg %>% left_join(pangene_stats) %>% left_join(sex)

```


```{r}
for (tissue_type in c("br", "e", "g", "h", "l", "bl")) {
  plot <- tpm_pg %>% filter(tissue==tissue_type, !is.na(chrom)) %>% 
    ggplot(aes(x=tpm, color=indiv)) + geom_freqpoly(bins=50) + ggtitle(tissue_type)
  print(plot)
}
```

Reference expression by variation outside reference.

```{r}

print("==== TISSUE: L ====")

tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "l") %>% select(gene_name, AW_var, tpm) %>% 
  lm(tpm ~ AW_var, data = .) %>% summary()

tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "l") %>% select(gene_name, AW_var, tpm) %>% 
  mutate(cnv = ifelse(AW_var > 1, "yes", "no")) %>% 
  t.test(tpm ~ cnv, conf.int = TRUE, data = .)


print("==== TISSUE:E ====")

tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "e") %>% select(gene_name, AW_var, tpm) %>% 
  lm(tpm ~ AW_var, data = .) %>% summary()

tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "e") %>% select(gene_name, AW_var, tpm) %>% 
  mutate(cnv = ifelse(AW_var > 1, "yes", "no")) %>% 
  t.test(tpm ~ cnv, conf.int = TRUE, data = .)

```

```{r}
tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "l") %>% select(gene_name, AW_var, tpm) %>% 
  group_by(gene_name, AW_var) %>%
  summarize(median_tpm = median(tpm, na.rm=TRUE)) %>% 
  mutate(AW_var = ifelse(AW_var > 5, "6+", as.character(AW_var))) %>%
  ggplot(aes(x=factor(AW_var, levels=c("0", "1", "2", "3", "4", "5", "6+")), y=median_tpm)) + geom_boxplot(outliers=FALSE, notch=TRUE) + theme_minimal() +
  labs(x = "Number of non-ref AW individuals", y = "Median log10(TPM) in Tissue: l")
```

```{r}
tpm_pg %>% filter(indiv=="AW_365336", count == 2, chrom!="non-aut", tissue == "e") %>% select(gene_name, AW_var, tpm) %>% 
  group_by(gene_name, AW_var) %>%
  summarize(median_tpm = median(tpm, na.rm=TRUE)) %>% 
  mutate(AW_var = ifelse(AW_var > 5, "6+", as.character(AW_var))) %>%
  ggplot(aes(x=factor(AW_var, levels=c("0", "1", "2", "3", "4", "5", "6+")), y=median_tpm)) + geom_boxplot(outliers=FALSE, notch=TRUE) + theme_minimal() +
  labs(x = "Number of non-ref AW individuals", y = "Median log10(TPM) in Tissue: e")

```

Now, let's look at expression variation by count. 


```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "l", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm, fill = sex)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "l", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + sex + (1|gene_name), data=.) %>% summary()

```

```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "l", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "l", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```




```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm, fill = sex)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + sex + (1|gene_name), data=.) %>% summary()

```

```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```




```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "h", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm, fill = sex)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "h", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + sex + (1|gene_name), data=.) %>% summary()

```

```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "h", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "h", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```



```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "br", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm, fill = sex)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "br", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + sex + (1|gene_name), data=.) %>% summary()

```

```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "br", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "br", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```



```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm, fill = sex)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + sex + (1|gene_name), data=.) %>% summary()

```

```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "e", ref_count > 0, AW_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()

```



```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "g", ref_count > 0, AW_var > 0, sex=="female") %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: g, female")

tpm_pg %>% filter(chrom != "non-aut", tissue == "g", ref_count > 0, AW_var > 0, sex=="male") %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```



```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "g", ref_count > 0, AW_var > 0, sex=="male") %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: g, male")

tpm_pg %>% filter(chrom != "non-aut", tissue == "g", ref_count > 0, AW_var > 0, sex=="male") %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```



```{r}
tpm_pg %>% filter(chrom != "non-aut", tissue == "bl", ref_count > 0, AC_var > 0) %>% 
  select(gene_name, indiv, count, tpm, sex) %>%
  ggplot(aes(x=as.factor(count), y=tpm)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: e")

tpm_pg %>% filter(chrom != "non-aut", tissue == "bl", ref_count > 0, AC_var > 0) %>% 
  select(gene_name, indiv, count, tpm) %>% distinct() %>% lmer(tpm ~ count + (1|gene_name), data=.) %>% summary()
```

All AW tissues together

```{r}
tpm_pg %>% filter(chrom != "non-aut", AW_var > 0, str_detect(indiv, "AW_")) %>% 
  select(gene_name, indiv, count, tpm, tissue, sex) %>% ungroup() %>%
  mutate(tissue = ifelse(tissue == "g" & sex == "female", "g.f", tissue)) %>%
  mutate(tissue = replace(tissue, tissue=="g", "g.m")) %>% 
  ggplot(aes(x=as.factor(count), y=tpm, fill=tissue)) + geom_boxplot(outliers=FALSE) + ggtitle("Tissue: all") + theme_minimal() + xlab("Copy Number") + ylab("TPM (log10)")

tpm_pg %>% filter(chrom != "non-aut", AW_var > 0, str_detect(indiv, "AW_")) %>% 
  select(gene_name, indiv, count, tpm, tissue, sex) %>% ungroup() %>%
  mutate(tissue = ifelse(tissue == "g" & sex == "female", "g.f", tissue)) %>%
  mutate(tissue = replace(tissue, tissue=="g", "g.m")) %>% 
  lmer(tpm ~ count + (1|gene_name) + (1|tissue), data=.) %>% summary()
```

